Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-11-01 text/supp.tex
Major edits to ms
After meeting with Nick, also reorder and incorporate
the supplementary figs

\item We removed all populations with uneven sampling intervals. I.e.~we removed
  populations that didn't have a constant difference between the ``decimal year
  begin'' and ``decimal year end'' columns.

@@

\item Populations with negative abundance values were assumed to be log values
  and transformed by taking $10$ to the power of the recorded abundance. In many
  cases this was noted for the population, but not in all cases. We inspected
  each of these \totalAssumedLog~time series to make sure our assumption made
  sense.

@@

\item We filled in single recorded values of zero with the lowest non-zero value
  in the time series \citep[following][]{brook2006a}. This assumes that single
  values of zero result from abundance being low enough that censusing missed
  present individuals. We turned multiple zero values in a row into \texttt{NA}
  values. This implies that multiple zero values were either censusing errors or
  caused by emigration. Regardless, our population models were fitted on
  a multiplicative (log) scale and so could not account for zero abundance. To
  avoid distorting the original data too strongly, we removed populations in
  which we filled in more than four zeros.

@@

Our prior on $\nu$ was based on \citet{fernandez1998}. They chose an
exponential rate parameter of $0.1$. We chose a less informative rate
parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$
the variance of the t distribution is undefined. This prior gives only
a $7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently
to avoid wandering off towards infinity --- above approximately $\nu = 20$ the
t distribution is so similar to the normal distribution
(Fig.~\ref{fig:didactic}) that time series of the length considered here are
unlikely to be informative about the precise value of $\nu$. In the scenario
where the data are informative about heavy tails
(e.g.~Fig.~\ref{fig:didactic}e,~h), the posterior will approximately match the
prior (prior median $= 71$, mean $= 102$) and the metrics used in our paper
(e.g.~Pr$(\nu < 10) > 0.5$) are unlikely to flag the population as heavy
tailed.

@@

We considered a version of the Gompertz model in which an autoregressive
parameter was fitted to the process noise residuals:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align*}
In addition to the parameters in the original Gompertz model, we estimate an
additional parameter $\phi$, which represents the dependence of
subsequent residuals. Based on the results of previous analyses with the GPDD
\citep[e.g.][]{connors2014} and the chosen priors in previous analyses
\citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when
running our model across all populations, we placed a weakly informative prior
on $\phi$ that assumed the greatest probability density near zero with the
reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose
$\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1,
\mathrm{max.} = 1)$. The MCMC chains did
not converge for \modelsNoConvergeAROne\ populations according to
our criteria ($\widehat{R} < 1.05, n_\mathrm{eff}
> 200$) after 8000 iterations of four chains. This included only
\modelsNoConvergeAROneHeavyBase\ populations in which
Pr($\nu < 10$) $> 0.5$ categorized them as heavy in the main Gompertz model.
We did not include these models in Fig.~\ref{fig:alt}.

@@

We fit a simplified version of the Gompertz model in which the density
dependence parameter $b$ was fixed at $1$ (density independent). This is
equivalent to fitting a random walk model (with drift) to the $\log$ abundances
or assuming the growth rates are drawn from a stationary distribution. The
model was as follows:
\begin{align*}
x_t &= \lambda + x_{t-1} + \epsilon_t\\
\epsilon &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align*}
We fit this model for three reasons: (1) it is computationally simpler and so
provides a check that our more complicated full Gompertz model was obtaining
ballpark reasonable estimates of $\nu$, (2) it provides a test of whether
density dependence was systematically affecting our perception of heavy tails,
(3) it matches how some previous authors have modelled heavy tails without
accounting for density dependence \citep{segura2013}.

@@

Observation error can bias parameter estimates \citep[e.g.][]{knape2012} and is
known to affect the ability to detect extreme events \citep{ward2007}. However,
simultaneously estimating observation and process noise is a challenging
problem and is known to result in identifiability issues with the Gompertz
population model \citep{knape2008}. Furthermore, our model estimates an
additional parameter --- the shape of the process deviation tails ---
potentially making identifiability and computational issues even greater.
Therefore, we considered a version of the base Gompertz model that allowed for
a fixed level of observation error:
\begin{align*}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
\end{align*}
where $U$ represents the unobserved state vector, $\sigma_\mathrm{obs}$
represents the standard deviation of observation error (on a log scale), and
$\sigma_\mathrm{proc}$ represent the process noise scale parameter. We set
$\sigma_\mathrm{obs}$ to $0.2$, which represents the upper limit of values often
used in simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.

@@

In the second part of our simulation testing, we tested the ability of the
heavy-tailed Gompertz model to obtain unbiased parameter estimates given that
process deviations were provided in which our simpler model (previous section)
could accurately estimate the value of $\nu$. To generate these process
deviations, we repeatedly drew proposed candidate process deviations and
estimated the central tendency, scale, and $\nu$ values each time. We recorded
when $\hat{\nu}$ (median of the posterior) was within $0.2$ CVs (coefficient of
variations) of the true $\nu$ value and used this set of random seed values in
our Gompertz simulation. We then fitted our Gompertz models to the simulated
datasets with all parameters (except $\nu$) set near the median values
estimated in the GPDD. We repeated this with $50$ and $100$ samples without
observation error, $50$ samples with observation error ($\sigma_\mathrm{obs}
= 0.2$), and $50$ samples with the same observation error and a Gompertz model
that allowed for correctly specified observation error magnitude.

@@

We fitted a multilevel beta regression model to the predicted probability of
heavy tails, Pr($\nu < 10$), to investigate potential covariates of heavy-tailed
dynamics. The beta distribution is useful when response data range on
a continuous scale between zero and one. We used a logit link function as is
typically used in logistic regression. The model was as follows:
\begin{align*}
\mathrm{Pr}(\nu_i < 0.5) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + X_i \beta),
  \: \text{for } i = 1, \dots, 617\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,\\
\end{align*}
where $A$ and $B$ represent the beta distribution shape parameters; $\mu_i$
represents the predicted value for population $i$, class $j$, order $k$, and
species $l$; $\phi_\mathrm{disp}$ represents the dispersion parameter; and $X_i$
represents a vector of predictors (such as lifespan) for population $i$ with
associated $\beta$ coefficients. The intercepts are allowed to vary from the
overall intercept $\alpha$ by taxonomic class ($\alpha^\mathrm{class}_j$),
taxonomic order ($\alpha^\mathrm{order}_k$), and species
($\alpha^\mathrm{species}_l$) with standard deviations $\sigma_{\alpha \;
  \mathrm{class}}$, $\sigma_{\alpha \; \mathrm{order}}$, and $\sigma_{\alpha \;
  \mathrm{species}}$. Where possible, we also allowed for error distributions
around the predictors by incorporating the standard deviation of the posterior
samples for the Gompertz parameters $\lambda$, $b$, and $\log \sigma$ around the
mean point value as normal distributions (not shown in the above equation).

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz-boxplots.pdf}
\caption{
  The distribution of median estimates from the same simulation as described in
  Fig.~\ref{fig:sim-gompertz}. The boxes show the interquartile range. The
  whiskers extend to $1.5$ times the interquartile range and outliers are shown
  as dots. Note that these distributions represent the output from only 20
  simulation iterations, so there is some level of remaining sampling noise.
}
\label{fig:sim-gompertz-boxplots}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{../analysis/ts-gpdd-heavy-eg-log10-base.pdf}
\caption{
  Time series for populations with Pr($\nu < 10$) $>$ 0.5 using the base
  Gompertz population model. Panels are ordered by increasing p($\nu < 10$).
  Vertical axes are on a log10 scale. Colours indicate taxonomic classes. The
  labels on each panel indicate p($\nu < 10$) $>$ 0.5, the common name for the
  species, and the Global Population Dynamics Database ID number.
}
\label{fig:heavy-ts}
\end{center}
\end{figure}

@@

>>>>

\item We removed all populations with uneven sampling intervals, i.e.\ we removed
  populations that didn't have a constant difference between the ``decimal year
  begin'' and ``decimal year end'' columns.

@@

\item Populations with negative abundance values were assumed to be log10
  values and transformed by taking $10$ to the power of the recorded abundance.
  In many cases this was noted for the population, but not in all cases. We
  inspected each of these \totalAssumedLog~time series to make sure our
  assumption made sense.

@@

\item We filled in single recorded values of zero with the lowest non-zero
  value in the time series \citep[following][]{brook2006a}. This assumes that
  single values of zero result from abundance being low enough that censusing
  overlooked individuals that were actually present. We turned multiple zero
  values in a row into \texttt{NA} values. This implies that multiple zero
  values were either censusing errors or caused by emigration. Regardless, our
  population models were fitted on a multiplicative (log) scale and so could
  not account for zero abundance. To avoid distorting the original data too
  strongly, we removed populations in which we filled in more than four zeros.

@@

Our prior on $\nu$ was based on \citet{fernandez1998}. They chose an
exponential rate parameter of $0.1$. We chose a less informative rate parameter
of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$ the
variance of the t distribution is undefined. This prior gives only a $7.7$\%
probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
wandering off towards infinity --- above approximately $\nu = 20$ the
t distribution is so similar to the normal distribution
(Fig.~\ref{fig:didactic}) that time series of the length considered here are
unlikely to be informative about the precise value of $\nu$. In the scenario
where the data are uninformative about heavy tails
(e.g.~Fig.~\ref{fig:didactic}e,~h), the posterior will approximately match the
prior (prior median $= 71$, mean $= 102$) and the metrics used in our paper
(e.g.~Pr$(\nu < 10) > 0.5$) are unlikely to flag the population as heavy
tailed.

@@

We considered a version of the Gompertz model in which an autoregressive
parameter was fitted to the process noise residuals:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align*}
In addition to the parameters in the original Gompertz model, we estimate an
additional parameter $\phi$, which represents the dependence of
subsequent residuals. Based on the results of previous analyses with the GPDD
\citep[e.g.][]{connors2014} and the chosen priors in previous analyses
\citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when
running our model across all populations, we placed a weakly informative prior
on $\phi$ that assumed the greatest probability density near zero with the
reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose
$\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -1,
\mathrm{max.} = 1)$. The MCMC chains did
not converge for \modelsNoConvergeAROne\ populations according to
our criteria ($\widehat{R} < 1.05, n_\mathrm{eff}
> 200$) after 8000 iterations of four chains. This included only
\modelsNoConvergeAROneHeavyBase\ populations in which
Pr($\nu < 10$) $> 0.5$ categorized them as heavy in the main Gompertz model.
We did not include these models in Fig.~\ref{fig:alt}.
\textit{TODO perhaps I should go up to 16,000 so that more converge.}

@@

We fit a simplified version of the Gompertz model in which the density
dependence parameter $b$ was fixed at $1$ (density independent). This is
equivalent to fitting a random walk model (with drift) to the $\log$ abundances
or assuming the growth rates are drawn from a stationary distribution. The
model was as follows:
\begin{align*}
x_t &= \lambda + x_{t-1} + \epsilon_t\\
\epsilon &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align*}
We fit this model for three reasons: (1) it is computationally simpler and so
provides a check that our more complicated full Gompertz model was obtaining
reasonable estimates of $\nu$, (2) it provides a test of whether density
dependence was systematically affecting our perception of heavy tails, (3) it
matches how some previous authors have modelled heavy tails without accounting
for density dependence \citep{segura2013}.

@@

Observation error can bias parameter estimates \citep[e.g.][]{knape2012} and is
known to affect the ability to detect extreme events \citep{ward2007}. In our
main analysis, we fit a model that ignored observation error. One way to
account for observation error would be to fit a full state-space model that
simultaneously estimates the magnitude of process noise and observation error.
However, simultaneously estimating observation and process noise is
a challenging problem (e.g.\ because the observation and process noise
parameters tend to negatively covary in model fitting) and is known to
sometimes result in identifiability issues with the Gompertz population model
\citep{knape2008}. Furthermore, our model was applied to hundreds of time
series, often of short length (as few as 20 time steps) and our model estimates
an additional parameter --- the shape of the process deviation tails ---
potentially making identifiability and computational issues even greater.
Therefore, we considered a version of the base Gompertz model that allowed for
a fixed level of observation error:
\begin{align*}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
\end{align*}
where $U$ represents the unobserved state vector, $\sigma_\mathrm{obs}$
represents the standard deviation of observation error (on a log scale), and
$\sigma_\mathrm{proc}$ represent the process noise scale parameter. We set
$\sigma_\mathrm{obs}$ to $0.2$, which represents the upper limit of values often
used in simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.

@@

In the second part of our simulation testing, we tested the ability of the
heavy-tailed Gompertz model to obtain unbiased parameter estimates 
when the process noise was chosen so that appropriate tail events were present.
To generate these process deviations for the $\nu = 3$ and $\nu = 5$ scenarios, we
repeatedly drew proposed candidate process deviations and estimated the central
tendency, scale, and $\nu$ values each time. We recorded when $\hat{\nu}$
(median of the posterior) was within $0.2$ CVs (coefficient of variations) of
the true $\nu$ value and used this set of random seed values in our Gompertz
simulation. The following simplified R code illustrates this procedure 
(the actual code is available at \url{https://github.com/seananderson/heavy-tails}):

\begin{footnotesize}
\begin{verbatim}
get_effective_nu_seeds <- function(nu_true = 5, cv = 0.2, N = 50, seed_N = 20) {
  # nu_true: The true nu value
  # cv:      The permitted effective nu coefficient of variation
  # N:       The length of time series
  # seed_N:  The number of seed values to generate
  seeds <- numeric(length = seed_N)
  seed_value <- 0
  for (i in seq_len(seed_N)) {
    nu_close <- FALSE
    while (!nu_close) {
      seed_value <- seed_value + 1
      set.seed(i)
      y <- rt(N, df = nu_true)
      sm <- rstan::stan(... # fit the Stan model here
      med_nu_hat <- median(rstan::extract(sm, pars = "nu")[[1]])
      if (med_nu_hat > (nu_true - cv) & med_nu_hat < (nu_true + cv)) {
        nu_close <- TRUE
        seeds[i] <- seed_value
      }
    }
  }
  seeds
}
nu_3_seeds_N50 <- get_effective_nu_seeds(nu_true = 3)
nu_5_seeds_N50 <- get_effective_nu_seeds(nu_true = 5)
\end{verbatim}
\end{footnotesize}

We then fitted our Gompertz models to the simulated datasets with all
parameters (except $\nu$) set near the median values estimated in the GPDD. We
repeated this with $50$ and $100$ samples without observation error, $50$
samples with observation error ($\sigma_\mathrm{obs} = 0.2$), and $50$ samples
with the same observation error and a Gompertz model that allowed for correctly
specified observation error magnitude.

@@

We fitted a multilevel beta regression model to the predicted probability of
heavy tails, Pr($\nu < 10$), to investigate potential covariates of heavy-tailed
dynamics. The beta distribution is useful when response data range on
a continuous scale between zero and one. We used a logit link function as is
typically used in logistic regression. The model was as follows:
\begin{align*}
\mathrm{Pr}(\nu_i < 0.5) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + X_i \beta),
  \: \text{for } i = 1, \dots, 617\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,
\end{align*}
where $A$ and $B$ represent the beta distribution shape parameters; $\mu_i$
represents the predicted value for population $i$, class $j$, order $k$, and
species $l$; $\phi_\mathrm{disp}$ represents the dispersion parameter; and $X_i$
represents a vector of predictors (such as lifespan) for population $i$ with
associated $\beta$ coefficients. The intercepts are allowed to vary from the
overall intercept $\alpha$ by taxonomic class ($\alpha^\mathrm{class}_j$),
taxonomic order ($\alpha^\mathrm{order}_k$), and species
($\alpha^\mathrm{species}_l$) with standard deviations $\sigma_{\alpha \;
  \mathrm{class}}$, $\sigma_{\alpha \; \mathrm{order}}$, and $\sigma_{\alpha \;
  \mathrm{species}}$. Where possible, we also allowed for error distributions
around the predictors by incorporating the standard deviation of the posterior
samples for the Gompertz parameters $\lambda$, $b$, and $\log \sigma$ around the
mean point value as normal distributions (not shown in the above equation).

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{../analysis/ts-gpdd-heavy-eg-log10-base.pdf}
\caption{
  Time series for populations with Pr($\nu < 10$) $>$ 0.5 using the base
  Gompertz population model. Panels are ordered by increasing Pr($\nu < 10$).
  Vertical axes are on a log10 scale. Colours indicate taxonomic classes. The
  labels on each panel indicate Pr($\nu < 10$) $>$ 0.5, the common name for the
  species, and the Global Population Dynamics Database ID number.
}
\label{fig:heavy-ts}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz-boxplots.pdf}
\caption{
  The distribution of median estimates from the same simulation as described in
  Fig.~\ref{fig:sim-gompertz}. The boxes show the interquartile range. The
  whiskers extend to $1.5$ times the interquartile range and outliers are shown
  as dots. Note that these distributions represent the output from only 20
  simulation iterations, so there is some level of remaining sampling noise.
  \textit{TODO wondering now if I should run these for more iterations to make
    these less noisy.}
}
\label{fig:sim-gompertz-boxplots}
\end{center}
\end{figure}
